# T-cells

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, cache=FALSE}
library(ggraph)
library(patchwork)
# CRC - Tcells
tcells_before_elbowDF = read.csv('data/tcells_before_elbowDF.csv', row.names=1)
tcells_before_meta = read.csv('data/tcells_before_meta.csv', row.names=1)
tcells_before_clustree = readRDS('data/tcells_before_clustree.Rdata')
SIGMA_all_res02 = readRDS('data/tcells_before_SIGMA.Rdata')
# CRC - Tcells_07
tcells_07_elbowDF = read.csv('data/tcells_07_elbowDF.csv', row.names=1)
tcells_07_meta = read.csv('data/tcells_07_meta.csv', row.names=1)
tcells_07_clustree = readRDS('data/tcells_07_clustree.Rdata')
# CRC - Tcells_1
tcells_1_elbowDF = read.csv('data/tcells_1_elbowDF.csv', row.names=1)
tcells_1_meta = read.csv('data/tcells_1_meta.csv', row.names=1)
tcells_1_clustree = readRDS('data/tcells_1_clustree.Rdata')
# CRC - Tcells_2
tcells_2_elbowDF = read.csv('data/tcells_2_elbowDF.csv', row.names=1)
tcells_2_meta = read.csv('data/tcells_2_meta.csv', row.names=1)
tcells_2_clustree = readRDS('data/tcells_2_clustree.Rdata')
```

To annotate the group of Tcells, we started out by loading the seurat object of Tcells

```{r eval=FALSE}
Tcells = SeuratDisk::LoadH5Seurat(paste(project_dir, '2_annotation/results_Tcells/datasets/Tcells.h5Seurat', sep='/'))
```


## First groups of T-cells

### Find Clusters

To find the clusters of T-cells, we started by performing a PCA. Scale was not performed, as the data was already scaled previously.

```{r eval=FALSE}
Tcells = Seurat::RunPCA(Tcells, assay='integrated')
invisible(gc())
Seurat::DefaultAssay(Tcells)='integrated'
```

Then, we calculated the number of PCs to use when finding the clusters in the next steps, by finding where the principal components start to elbow.

```{r eval=FALSE}
# Calculate the first metric:
pct = Tcells[["pca"]]@stdev /sum(Tcells[["pca"]]@stdev) * 100
cumu = cumsum(pct)
co1 = which(cumu > 90 & pct < 5)[1]

# Calculate the second metric:
co2 = sort(which((pct[1:length(pct) - 1] - pct[2:length(pct)]) > 0.1), decreasing = T)[1] + 1

# Where the elbow occurs:
elbow = min(co1, co2) # elbow is 15
```

This is a visual representation of the elbow plot:

```{r eval=FALSE}
plot_df = data.frame(pct=pct, cumu=cumu, rank=1:length(pct))
ggplot2::ggplot(plot_df, ggplot2::aes(cumu, pct, label = rank, color = rank > elbow)) + 
  ggplot2::geom_text() + 
  ggplot2::geom_vline(xintercept = 90, color = "grey") + 
  ggplot2::geom_hline(yintercept = min(pct[pct > 5]), color = "grey") +
  ggplot2::theme_bw() + ggplot2::theme(legend.position = "bottom")
```

```{r fig.width=10, fig.height=5, out.width='80%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.cap="Elbow plot. The last PC (15) colored in red corresponds to the elbow.", fig.align='center'}
elbow=15
ggplot2::ggplot(data=tcells_before_elbowDF, ggplot2::aes_string('cumu', 'pct', label = 'rank', color = tcells_before_elbowDF$rank > elbow)) + 
  ggplot2::geom_text() + 
  ggplot2::geom_vline(xintercept = 90, color = "grey") + 
  ggplot2::geom_hline(yintercept = min(tcells_before_elbowDF$pct[tcells_before_elbowDF$pct > 5]), color = "grey") +
  ggplot2::theme_bw() + ggplot2::theme(legend.position = "none")
```

After finding the elbow, we determined the K-nearest neighbors using the calculated elbow and found the clusters. We used a set of 10 different resolutions, ranging from 0.1 to 1 at a step of 0.1.

```{r eval=FALSE}
# Determine the K-nearest neighbor graph
Tcells = Seurat::FindNeighbors(Tcells, dims=1:elbow)

# Find clusters:
Tcells = Seurat::FindClusters(Tcells, resolution=seq(0.1, 1, by=.1))

# Run UMAP visualization:
Tcells = Seurat::RunUMAP(Tcells, dims=1:elbow, reduction='pca')
```


### Check Quality of Clusters

We explored if certain metrics and metadata are source of variation, which helps assessing the quality of the clustering

```{r eval=FALSE}
metrics =  c("nUMI", "nGene", "S.Score", "G2M.Score", "percent.mitochondrial_RNA")
feature_plots(Tcells, metrics, ncol=2, with_dimplot=F, point_size=0.2)
Seurat::DimPlot(Tcells, reduction="umap", group.by=c('dataset', 'Phase', 'state'), label=FALSE)
```

```{r fig.width=10, fig.height=20, out.width='80%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.align='center'}
list_plots = list()

list_plots[['nUMI']] =  ggplot2::ggplot(tcells_before_meta, ggplot2::aes(x=UMAP_1, y=UMAP_2, color=nUMI)) +
  ggplot2::geom_point(size=.5) + ggplot2::ggtitle("nUMI") +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) + ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

list_plots[['nGene']] =  ggplot2::ggplot(tcells_before_meta, ggplot2::aes(x=UMAP_1, y=UMAP_2, color=nGene)) +
  ggplot2::geom_point(size=.5) + ggplot2::ggtitle("nGene") +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) + ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

list_plots[['S.Score']] =  ggplot2::ggplot(tcells_before_meta, ggplot2::aes(x=UMAP_1, y=UMAP_2, color=S.Score)) +
  ggplot2::geom_point(size=.5) + ggplot2::ggtitle("S.Score") +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) + ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

list_plots[['G2M.Score']] =  ggplot2::ggplot(tcells_before_meta, ggplot2::aes(x=UMAP_1, y=UMAP_2, color=G2M.Score)) +
  ggplot2::geom_point(size=.5) + ggplot2::ggtitle("G2M.Score") +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) + ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

list_plots[['percent.mitochondrial_RNA']] =  ggplot2::ggplot(tcells_before_meta, ggplot2::aes(x=UMAP_1, y=UMAP_2, color=percent.mitochondrial_RNA)) +
  ggplot2::geom_point(size=.5) + ggplot2::ggtitle("percent.mitochondrial_RNA") +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) + ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

list_plots[['dataset']] =  ggplot2::ggplot(tcells_before_meta, ggplot2::aes(x=UMAP_1, y=UMAP_2, color=dataset)) +
  ggplot2::geom_point(size=.2) + ggplot2::ggtitle("dataset") +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

list_plots[['Phase']] =  ggplot2::ggplot(tcells_before_meta, ggplot2::aes(x=UMAP_1, y=UMAP_2, color=Phase)) +
  ggplot2::geom_point(size=.2) + ggplot2::ggtitle("Phase") +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

list_plots[['state']] =  ggplot2::ggplot(tcells_before_meta, ggplot2::aes(x=UMAP_1, y=UMAP_2, color=state)) +
  ggplot2::geom_point(size=.2) + ggplot2::ggtitle("state") +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")


patchwork::wrap_plots(list_plots, ncol=2)
```


### Choose Cluster Resolution

#### Check clustree results and gene expression

We used 10 different resolutions so that we could find the resolution that best separates the cells into clusters with interesting biological information.The best resolution was chosen based on the stability of the clusters between different resolutions and the expression of cell-type markers, with the help of the [clustree](https://cran.r-project.org/web/packages/clustree/index.html) package [@R-clustree] (v.0.4.3).

The *clustree* results, accompanied with UMAP visualizations for 4 different resolutions, is shown:

```{r eval=FALSE}
# 4. Use clustree:
library(ggraph)
clust_tree = clustree::clustree(Tcells, prefix='integrated_snn_res.')
# Visualize 4 different resolutions:
clusters_01 = Seurat::DimPlot(Tcells, reduction="umap", group.by='integrated_snn_res.0.1', label=TRUE, label.size=6, pt.size=.2)
clusters_04 = Seurat::DimPlot(Tcells, reduction="umap", group.by='integrated_snn_res.0.2', label=TRUE, label.size=6, pt.size=.2)
clusters_07 = Seurat::DimPlot(Tcells, reduction="umap", group.by='integrated_snn_res.0.3', label=TRUE, label.size=6, pt.size=.2)
clusters_08 = Seurat::DimPlot(Tcells, reduction="umap", group.by='integrated_snn_res.0.4', label=TRUE, label.size=6, pt.size=.2)
((clusters_01 | clusters_02) / (clusters_03 | clusters_04)) | clust_tree
```

```{r fig.width=15, fig.height=10, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
clusters = list()
for(res in c('integrated_snn_res.0.1', 'integrated_snn_res.0.2', 'integrated_snn_res.0.3', 'integrated_snn_res.0.4')){
  tcells_before_meta[,res] = factor(as.character(tcells_before_meta[,res]), levels=as.character(0:max(tcells_before_meta[,res])))
  x = c()
  y = c()
  for(cluster in levels(tcells_before_meta[,res])){
    x = c(x, mean(tcells_before_meta$UMAP_1[as.character(tcells_before_meta[,res])==cluster]))
    y = c(y, mean(tcells_before_meta$UMAP_2[as.character(tcells_before_meta[,res])==cluster]))
  }
  clusters[[res]] = ggplot2::ggplot(tcells_before_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=res)) +
    ggplot2::geom_point(size=.2) + ggplot2::ggtitle(res) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_before_meta[,res]), color='black')
}
((clusters$integrated_snn_res.0.1 | clusters$integrated_snn_res.0.2) / (clusters$integrated_snn_res.0.3 | clusters$integrated_snn_res.0.4)) | tcells_before_clustree
```

We next assessed the expression of T-cell gene markers mapped into UMAP plots and violin plots (this last type of plots with resolution 0.2 as reference):


| Genes            | Cell-Types            |
|:---------------- |:--------------------- |
| CD3E, CD3D, CD3G | T-cells               |
| CD8A             | CD8+ T-cells          |
| CD4              | CD4+ T-cells          |
| TRGC1, TRGC2     | $\gamma\delta$T-cells |
| KLRB1            | NK-like               |
| CXCL13           | CXCL13+ T-cells       |
| IL2RA            | Regulatory T-cells    |
| IL17A            | Th17 cells            |

```{r eval=FALSE}
umap = Seurat::FeaturePlot(Tcells, reduction='umap', features=c('rna_CD3E', 'rna_CD3D', 'rna_CD3G', 'rna_CD8A', 'rna_CD4', 'rna_TRGC1', 'rna_TRGC2', 'rna_KLRB1', 'rna_CXCL13', 'rna_IL2RA', 'rna_IL17A'), label=TRUE, ncol=1) +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

vln = Seurat::VlnPlot(Tcells, c('rna_CD3E', 'rna_CD3D', 'rna_CD3G', 'rna_CD8A', 'rna_CD4', 'rna_TRGC1', 'rna_TRGC2', 'rna_KLRB1', 'rna_CXCL13', 'rna_IL2RA', 'rna_IL17A'), ncol=1) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none")

umap | vln
```

```{r fig.width=10, fig.height=20, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.align='center'}
x = c()
y = c()
for(cluster in levels(tcells_before_meta$integrated_snn_res.0.2)){
  x = c(x, mean(tcells_before_meta$UMAP_1[as.character(tcells_before_meta$integrated_snn_res.0.2)==cluster]))
  y = c(y, mean(tcells_before_meta$UMAP_2[as.character(tcells_before_meta$integrated_snn_res.0.2)==cluster]))
}

umaps = list()
vlns = list()

for(gene in c('rna_CD3E', 'rna_CD3D', 'rna_CD3G', 'rna_CD8A', 'rna_CD4', 'rna_TRDC', 'rna_TRGC1', 'rna_TRGC2', 'rna_KLRB1', 'rna_CXCL13', 'rna_IL2RA', 'rna_IL17A')){
  umaps[[gene]] = ggplot2::ggplot(tcells_before_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=gene)) +
    ggplot2::geom_point(size=.6) + ggplot2::ggtitle(gsub('rna_', '', gene)) +
    ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2),
                                   limits=c(1e-5, NA)) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_before_meta$integrated_snn_res.0.2), color='black')
  
  vlns[[gene]] = ggplot2::ggplot(tcells_before_meta, ggplot2::aes_string(x='integrated_snn_res.0.2', y=gene,
                                                                   fill='integrated_snn_res.0.2')) +
    ggplot2::geom_violin(trim=FALSE, scale='width') + ggplot2::theme_minimal() + ggplot2::theme(legend.position="none") +
    ggplot2::labs(x='', y='', title=gsub('rna_', '', gene)) + ggplot2::ylim(c(0,max(tcells_before_meta[, gene])))
}

final_plots = list()
for(gene in c('rna_CD3E', 'rna_CD3D', 'rna_CD3G', 'rna_CD8A', 'rna_CD4', 'rna_TRDC', 'rna_TRGC1', 'rna_TRGC2', 'rna_KLRB1', 'rna_CXCL13', 'rna_IL2RA', 'rna_IL17A')){
  final_plots[[paste('a', gene, sep='_')]] = umaps[[gene]]
  final_plots[[paste('b', gene, sep='_')]] = vlns[[gene]]
}
patchwork::wrap_plots(final_plots, ncol=4)
```

#### Subset T-cells for further sub-clustering

After looking into the expression of these T-cell gene markers (and others, not shown) mapped into UMAP plots and violin plots (for resolution 0.2), as well as the clustree results, we decided to assess if it made sense to separate the clusters according to resolution 0.2 and further sub-cluster each of these individually. Our main objective was to separate T-cells not only into CD4+, CD8+, regulatory, NK and non-conventional T-cells, but into further sub-clusters of these cell-types (which was not possible with the same pipeline applied to the previous groups of cells).

Looking into the resolution 0.2, we could know at this point that the clusters corresponded, more or less, to the following major annotations:

| Clusters | Annotations                        |
|:-------- |:---------------------------------- |
| 0        | CD4+                               |
| 1        | CD8+                               |
| 2        | Regulatory Tcells                  |
| 3        | gd Tcells + NK-like + ILCs         |
| 4        | CD8+ + CD4+ expressing CXCL13      |
| 5        | CD4+ (mostly Th17)                 |
| 6        | Proliferative T-cells              |
| 7        | CD4+                               |

We next used the [SIGMA](https://github.com/Siliegia/SIGMA) package [@R-SIGMA] (v.0.0.0.1) to check how clusterable each cluster in resolution 0.2 were, and if that was due to meaningful variability (for example, we do not want to end up clustering cells according to dataset of origin).

We first subsetted our *Tcells* dataset to a quarter of the original number of cells, for computational reasons. We used the integrated counts, as those were already corrected for dataset of origin, cell cycle and percentage of mitochondrial RNA.

```{r eval=FALSE}
# 1. Subset cells to a quarter of the original
n_cells = round(dim(Tcells@assays$integrated@data)[2] / 4)
sampled_cells = colnames(Tcells@assays$integrated@data)[sample(1:dim(Tcells@assays$integrated@data)[2], n_cells)]

# 2. Check if subsetting still resembles unsubsetted dataset
clusters_02 | Seurat::DimPlot(subset(Tcells, cells=sampled_cells), reduction="umap", group.by='integrated_snn_res.0.2',
                              label=TRUE, label.size=6, pt.size=.2)

# 3. Subset dataset
Tcells_subset = subset(Tcells, cells=sampled_cells)
invisible(gc())

# 4. Get integrated dataset
sigma_data = as.matrix(SeuratObject::GetAssayData(Tcells_subset, assay='integrated', slot='data'))
invisible(gc())

# 5. Get vector with resolution 0.2 clusters and dataset
clusters_02_vector = Tcells_subset@meta.data[colnames(sigma_data), 'integrated_snn_res.0.2']
dataset_vector = Tcells_subset@meta.data[colnames(sigma_data), 'dataset']
```

After successfully processing the data for *SIGMA*, we ran the tool.

```{r eval=FALSE}
SIGMA_all_res02 = SIGMA::sigma_funct(sigma_data, clusters_02_vector)

# 2.1. Save SIGMA result
saveRDS(SIGMA_all_res02, paste(project_dir, '2_annotation/results_Tcells/datasets/SIGMA_all_res02.Rdata', sep='/'))
```

After obtaining the results, we first checked the clusterability for all clusters. This is measured by the metric *SIGMA*, which goes from 0 to 1. The closer to 1, the more clusterable.

```{r fig.width=10, fig.height=5, out.width='80%', fig.align='center', warning=FALSE, error=FALSE, message=FALSE, results='hold'}
SIGMA::plot_sigma(SIGMA_all_res02)

SIGMA_all_res02$maximum_measure
```

Then, we visually checked if the clusterability was not due to dataset of origin:

```{r eval=FALSE}
dtSIGMA_C0 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '0', colour=dataset_vector[Tcells_subset@meta.data$integrated_snn_res.0.2=='0']) + ggplot2::ggtitle('Cluster 0')

dtSIGMA_C1 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '1', colour=dataset_vector[Tcells_subset@meta.data$integrated_snn_res.0.2=='1']) + ggplot2::ggtitle('Cluster 1')

dtSIGMA_C2 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '2', colour=dataset_vector[Tcells_subset@meta.data$integrated_snn_res.0.2=='2']) + ggplot2::ggtitle('Cluster 2')

dtSIGMA_C3 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '3', colour=dataset_vector[Tcells_subset@meta.data$integrated_snn_res.0.2=='3']) + ggplot2::ggtitle('Cluster 3')

dtSIGMA_C4 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '4', colour=dataset_vector[Tcells_subset@meta.data$integrated_snn_res.0.2=='4']) + ggplot2::ggtitle('Cluster 4')

dtSIGMA_C5 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '5', colour=dataset_vector[Tcells_subset@meta.data$integrated_snn_res.0.2=='5']) + ggplot2::ggtitle('Cluster 5')

dtSIGMA_C6 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '6', colour=dataset_vector[Tcells_subset@meta.data$integrated_snn_res.0.2=='6']) + ggplot2::ggtitle('Cluster 6')

dtSIGMA_C7 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '7', colour=dataset_vector[Tcells_subset@meta.data$integrated_snn_res.0.2=='7']) + ggplot2::ggtitle('Cluster 7')

patchwork::wrap_plots(list(dtSIGMA_C0, dtSIGMA_C1, dtSIGMA_C2, dtSIGMA_C3, dtSIGMA_C4, dtSIGMA_C5, dtSIGMA_C6,dtSIGMA_C7), ncol=2)
```

```{r results='hide', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, include=FALSE}
dataset_vector = tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr), 'dataset']
tcells_before_meta$integrated_snn_res.0.2 = factor(as.character(tcells_before_meta$integrated_snn_res.0.2), as.character(0:(length(unique(tcells_before_meta$integrated_snn_res.0.2))-1)))

dtSIGMA_C0 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '0', colour=dataset_vector[tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr),'integrated_snn_res.0.2']=='0']) + ggplot2::ggtitle('Cluster 0')

dtSIGMA_C1 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '1', colour=dataset_vector[tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr),'integrated_snn_res.0.2']=='1']) + ggplot2::ggtitle('Cluster 1')

dtSIGMA_C2 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '2', colour=dataset_vector[tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr),'integrated_snn_res.0.2']=='2']) + ggplot2::ggtitle('Cluster 2')

dtSIGMA_C3 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '3', colour=dataset_vector[tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr),'integrated_snn_res.0.2']=='3']) + ggplot2::ggtitle('Cluster 3')

dtSIGMA_C4 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '4', colour=dataset_vector[tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr),'integrated_snn_res.0.2']=='4']) + ggplot2::ggtitle('Cluster 4')

dtSIGMA_C5 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '5', colour=dataset_vector[tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr),'integrated_snn_res.0.2']=='5']) + ggplot2::ggtitle('Cluster 5')

dtSIGMA_C6 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '6', colour=dataset_vector[tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr),'integrated_snn_res.0.2']=='6']) + ggplot2::ggtitle('Cluster 6')

dtSIGMA_C7 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '7', colour=dataset_vector[tcells_before_meta[colnames(SIGMA_all_res02$input_parameters$expr),'integrated_snn_res.0.2']=='7']) + ggplot2::ggtitle('Cluster 7')

sigma_plots = patchwork::wrap_plots(list(dtSIGMA_C0, dtSIGMA_C1, dtSIGMA_C2, dtSIGMA_C3, dtSIGMA_C4, dtSIGMA_C5, dtSIGMA_C6,dtSIGMA_C7), ncol=2)
```

```{r fig.width=15, fig.height=20, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
sigma_plots
```

With the exception of cluster 7, which had no clusterability (*SIGMA* = 0), all other clusters had a clusterability higher than 0.89. Also, the variability in the clusters with high *SIGMA* values was not coming from the source dataset.

With these results, we decided we could subcluster each cluster individually to obtain more fine-grained annotations.


## Annotate Clusters 0 and 7

When looking into the UMAP visualizations of the Tcells clusters from resolution 0.2, including the expression of gene markers and how cluster 7 did not have any clusterability, clusters 0 and 7 seemed to represent the same cell-type(s). As such, we decided to merge these two clusters together.

```{r eval=FALSE}
Tcells_07 = subset(Tcells, cells=rownames(Tcells@meta.data)[Tcells$integrated_snn_res.0.2%in%c('0', '7')])
Tcells_07[['previous_0_7']] = Tcells_07$integrated_snn_res.0.2

(Seurat::DimPlot(Tcells, group.by = 'integrated_snn_res.0.2', label=T, label.size=4) + ggplot2::theme_minimal() + Seurat::NoLegend()) |
  (Seurat::DimPlot(Tcells_07, group.by = 'previous_0_7', label=T, label.size=4) + ggplot2::theme_minimal() + Seurat::NoLegend())
```

```{r fig.width=15, fig.height=5, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
cells_07 = rownames(tcells_before_meta)[tcells_before_meta$integrated_snn_res.0.2%in%c('0', '7')]

x = c()
y = c()
for(cluster in levels(tcells_before_meta[,'integrated_snn_res.0.2'])){
  x = c(x, mean(tcells_before_meta$UMAP_1[as.character(tcells_before_meta[,'integrated_snn_res.0.2'])==cluster]))
  y = c(y, mean(tcells_before_meta$UMAP_2[as.character(tcells_before_meta[,'integrated_snn_res.0.2'])==cluster]))
}
tcells_all = ggplot2::ggplot(tcells_before_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color='integrated_snn_res.0.2')) +
  ggplot2::geom_point(size=.2) + ggplot2::ggtitle('All clusters') +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none") +
  ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_before_meta[,'integrated_snn_res.0.2']), color='black')

x2 = c()
y2 = c()
for(cluster in levels(tcells_before_meta[cells_07,'integrated_snn_res.0.2'])){
  x2 = c(x2, mean(tcells_before_meta$UMAP_1[as.character(tcells_before_meta[,'integrated_snn_res.0.2'])==cluster]))
  y2 = c(y2, mean(tcells_before_meta$UMAP_2[as.character(tcells_before_meta[,'integrated_snn_res.0.2'])==cluster]))
}
tcells_07 = ggplot2::ggplot(tcells_before_meta[cells_07,], ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color='integrated_snn_res.0.2')) +
  ggplot2::geom_point(size=.2) + ggplot2::ggtitle('Only clusters 0 and 7') +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none") +
  ggplot2::annotate(geom='text', x=x2, y=y2, label=levels(tcells_before_meta[,'integrated_snn_res.0.2']), color='black')

tcells_all | tcells_07
```


### Find Clusters

After subsetting the dataset to obtain only the cells from clusters 0 and 7, we prepared the data for clustering by running PCA and calculating the elbow.

```{r eval=FALSE}
# PCA:
Tcells_07 = Seurat::RunPCA(Tcells_07, assay='integrated')

# Calculate Elbow:
# First metric:
pct = Tcells_07[["pca"]]@stdev /sum(Tcells_07[["pca"]]@stdev) * 100
cumu = cumsum(pct)
co1 = which(cumu > 90 & pct < 5)[1]
# Second metric:
co2 = sort(which((pct[1:length(pct) - 1] - pct[2:length(pct)]) > 0.1), decreasing = T)[1] + 1
# Where the elbow occurs:
elbow = min(co1, co2) # elbow is 5
```

This is a visual representation of the elbow plot:

```{r eval=FALSE}
plot_df = data.frame(pct=pct, cumu=cumu, rank=1:length(pct))
ggplot2::ggplot(plot_df, ggplot2::aes(cumu, pct, label = rank, color = rank > elbow)) + 
  ggplot2::geom_text() + 
  ggplot2::geom_vline(xintercept = 90, color = "grey") + 
  ggplot2::geom_hline(yintercept = min(pct[pct > 5]), color = "grey") +
  ggplot2::theme_bw() + ggplot2::theme(legend.position = "bottom")
```

```{r fig.width=10, fig.height=5, out.width='80%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.cap="Elbow plot. The last PC (5) colored in red corresponds to the elbow.", fig.align='center'}
elbow=5
ggplot2::ggplot(data=tcells_07_elbowDF, ggplot2::aes_string('cumu', 'pct', label = 'rank', color = tcells_07_elbowDF$rank > elbow)) + 
  ggplot2::geom_text() + 
  ggplot2::geom_vline(xintercept = 90, color = "grey") + 
  ggplot2::geom_hline(yintercept = min(tcells_07_elbowDF$pct[tcells_07_elbowDF$pct > 5]), color = "grey") +
  ggplot2::theme_bw() + ggplot2::theme(legend.position = "none")
```

After finding the elbow, we determined the K-nearest neighbors using the calculated elbow and found the clusters. We also used a set of 10 different resolutions, ranging from 0.1 to 1 at a step of 0.1.

```{r eval=FALSE}
# Determine the K-nearest neighbor graph
Tcells_07 = Seurat::FindNeighbors(Tcells_07, dims=1:elbow)

# Find clusters:
Tcells_07 = Seurat::FindClusters(Tcells_07, resolution=seq(0.1, 1, by=.1))

# Run UMAP visualization:
Tcells_07 = Seurat::RunUMAP(Tcells_07, dims=1:elbow, reduction='pca')
```

### Choose Cluster Resolution

Again, we used 10 different resolutions so that we could find the resolution that best separates the cells into clusters with interesting biological information.The best resolution was chosen based on the stability of the clusters between different resolutions and the expression of cell-type markers, with the help of the [clustree](https://cran.r-project.org/web/packages/clustree/index.html) package [@R-clustree] (v.0.4.3).

The *clustree* results, accompanied with UMAP visualizations for 3 different resolutions, is shown. We also show a UMAP with the cells colored by the  'cluster of origin' (clusters 0 and 7 from full dataset of Tcells), to assess if they are being separated by these two clusters and not by cell-type.

```{r eval=FALSE}
# 4. Use clustree:
library(ggraph)
clust_tree = clustree::clustree(Tcells_07, prefix='integrated_snn_res.')
# Visualize 4 different resolutions:
clusters_01 = Seurat::DimPlot(Tcells_07, reduction="umap", group.by='integrated_snn_res.0.1', label=TRUE, label.size=6, pt.size=.2)
clusters_04 = Seurat::DimPlot(Tcells_07, reduction="umap", group.by='integrated_snn_res.0.2', label=TRUE, label.size=6, pt.size=.2)
clusters_07 = Seurat::DimPlot(Tcells_07, reduction="umap", group.by='integrated_snn_res.0.9', label=TRUE, label.size=6, pt.size=.2)
clusters_08 = Seurat::DimPlot(Tcells_07, reduction="umap", group.by='previous_0_7', label=TRUE, label.size=6, pt.size=.2)
((clusters_01 | clusters_04) / (clusters_07 | clusters_08)) | clust_tree
```

```{r fig.width=15, fig.height=10, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
clusters = list()
for(res in c('integrated_snn_res.0.1', 'integrated_snn_res.0.2', 'integrated_snn_res.0.9', 'previous_0_7')){
  tcells_07_meta[,res] = factor(as.character(tcells_07_meta[,res]), levels=as.character(0:max(tcells_07_meta[,res])))
  x = c()
  y = c()
  for(cluster in levels(tcells_07_meta[,res])){
    x = c(x, mean(tcells_07_meta$UMAP_1[as.character(tcells_07_meta[,res])==cluster]))
    y = c(y, mean(tcells_07_meta$UMAP_2[as.character(tcells_07_meta[,res])==cluster]))
  }
  clusters[[res]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=res)) +
    ggplot2::geom_point(size=.2) + ggplot2::ggtitle(res) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_07_meta[,res]), color='black')
}
((clusters$integrated_snn_res.0.1 | clusters$integrated_snn_res.0.2) / (clusters$integrated_snn_res.0.9 | clusters$previous_0_7)) | tcells_07_clustree
```

It is possible to see that 'cluster of origin' is not a source of variation between the clusters calculated. Furthermore, the resolution that we ended up choosing was 0.1, after looking into the *clustree* results above, and the expression of gene markers that we show next.

### Check Gene Expression

The following markers were used to annotate the cell-types:

| Genes                                     | Cell-Types                           |
|:----------------------------------------- |:------------------------------------ |
| CD3E, CD3D, CD3G, CD4, TRAC, TRBC1, TRBC2 | CD4+ T-cells                         |
| CCR7, SELL                                | Naive T-cells                        |
| CCR7, SELL, PASK                          | Central Memory (CM) T-cells          |
| ZNF683                                    | Tissue Resident Memory (TRM) T-cells |
| KLRB1, S100A4, PTPRCAP, PFN1, GZMA        | Effector Memory (EM) T-cells         |

To assess the expression of these marker genes, we visualized their expression mapped into UMAP plots and violin plots.

#### CD4+ T-cells

```{r eval=FALSE}
umap = Seurat::FeaturePlot(Tcells_07, reduction='umap', features=c('rna_CD3E', 'rna_CD3D', 'rna_CD3G', 'rna_CD4', 'rna_TRAC', 'rna_TRBC1', 'rna_TRBC2'), label=TRUE, ncol=1) +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

vln = Seurat::VlnPlot(Tcells_07, c('rna_CD3E', 'rna_CD3D', 'rna_CD3G', 'rna_CD4', 'rna_TRAC', 'rna_TRBC1', 'rna_TRBC2'), ncol=1) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none")

umap | vln
```

```{r fig.width=10, fig.height=15, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.align='center'}
x = c()
y = c()
for(cluster in levels(tcells_07_meta$integrated_snn_res.0.1)){
  x = c(x, mean(tcells_07_meta$UMAP_1[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
  y = c(y, mean(tcells_07_meta$UMAP_2[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
}

umaps = list()
vlns = list()

for(gene in c('rna_CD3E', 'rna_CD3D', 'rna_CD3G', 'rna_CD4', 'rna_TRAC', 'rna_TRBC1', 'rna_TRBC2')){
  umaps[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=gene)) +
    ggplot2::geom_point(size=.6) + ggplot2::ggtitle(gsub('rna_', '', gene)) +
    ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2),
                                   limits=c(1e-5, NA)) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_07_meta$integrated_snn_res.0.1), color='black')
  
  vlns[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='integrated_snn_res.0.1', y=gene,
                                                                   fill='integrated_snn_res.0.1')) +
    ggplot2::geom_violin(trim=FALSE, scale='width') + ggplot2::theme_minimal() + ggplot2::theme(legend.position="none") +
    ggplot2::labs(x='', y='', title=gsub('rna_', '', gene)) + ggplot2::ylim(c(0,max(tcells_07_meta[, gene])))
}

final_plots = list()
for(gene in c('rna_CD3E', 'rna_CD3D', 'rna_CD3G', 'rna_CD4', 'rna_TRAC', 'rna_TRBC1', 'rna_TRBC2')){
  final_plots[[paste('a', gene, sep='_')]] = umaps[[gene]]
  final_plots[[paste('b', gene, sep='_')]] = vlns[[gene]]
}
patchwork::wrap_plots(final_plots, ncol=4)
```

#### Naive T-cells

```{r eval=FALSE}
umap = Seurat::FeaturePlot(Tcells_07, reduction='umap', features=c('rna_CCR7', 'rna_SELL'), label=TRUE, ncol=1) +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

vln = Seurat::VlnPlot(Tcells_07, c('rna_CCR7', 'rna_SELL'), ncol=1) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none")

umap | vln
```

```{r fig.width=10, fig.height=5, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.align='center'}
x = c()
y = c()
for(cluster in levels(tcells_07_meta$integrated_snn_res.0.1)){
  x = c(x, mean(tcells_07_meta$UMAP_1[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
  y = c(y, mean(tcells_07_meta$UMAP_2[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
}

umaps = list()
vlns = list()

for(gene in c('rna_CCR7', 'rna_SELL')){
  umaps[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=gene)) +
    ggplot2::geom_point(size=.6) + ggplot2::ggtitle(gsub('rna_', '', gene)) +
    ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2),
                                   limits=c(1e-5, NA)) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_07_meta$integrated_snn_res.0.1), color='black')
  
  vlns[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='integrated_snn_res.0.1', y=gene,
                                                                   fill='integrated_snn_res.0.1')) +
    ggplot2::geom_violin(trim=FALSE, scale='width') + ggplot2::theme_minimal() + ggplot2::theme(legend.position="none") +
    ggplot2::labs(x='', y='', title=gsub('rna_', '', gene)) + ggplot2::ylim(c(0,max(tcells_07_meta[, gene])))
}

final_plots = list()
for(gene in c('rna_CCR7', 'rna_SELL')){
  final_plots[[paste('a', gene, sep='_')]] = umaps[[gene]]
  final_plots[[paste('b', gene, sep='_')]] = vlns[[gene]]
}
patchwork::wrap_plots(final_plots, ncol=4)
```

#### Central Memory (CM) T-cells

```{r eval=FALSE}
umap = Seurat::FeaturePlot(Tcells_07, reduction='umap', features=c('rna_CCR7', 'rna_SELL', 'rna_PASK'), label=TRUE, ncol=1) +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

vln = Seurat::VlnPlot(Tcells_07, c('rna_CCR7', 'rna_SELL', 'rna_PASK'), ncol=1) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none")

umap | vln
```

```{r fig.width=10, fig.height=10, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.align='center'}
x = c()
y = c()
for(cluster in levels(tcells_07_meta$integrated_snn_res.0.1)){
  x = c(x, mean(tcells_07_meta$UMAP_1[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
  y = c(y, mean(tcells_07_meta$UMAP_2[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
}

umaps = list()
vlns = list()

for(gene in c('rna_CCR7', 'rna_SELL', 'rna_PASK')){
  umaps[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=gene)) +
    ggplot2::geom_point(size=.6) + ggplot2::ggtitle(gsub('rna_', '', gene)) +
    ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2),
                                   limits=c(1e-5, NA)) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_07_meta$integrated_snn_res.0.1), color='black')
  
  vlns[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='integrated_snn_res.0.1', y=gene,
                                                                   fill='integrated_snn_res.0.1')) +
    ggplot2::geom_violin(trim=FALSE, scale='width') + ggplot2::theme_minimal() + ggplot2::theme(legend.position="none") +
    ggplot2::labs(x='', y='', title=gsub('rna_', '', gene)) + ggplot2::ylim(c(0,max(tcells_07_meta[, gene])))
}

final_plots = list()
for(gene in c('rna_CCR7', 'rna_SELL', 'rna_PASK')){
  final_plots[[paste('a', gene, sep='_')]] = umaps[[gene]]
  final_plots[[paste('b', gene, sep='_')]] = vlns[[gene]]
}
patchwork::wrap_plots(final_plots, ncol=4)
```

#### Tissue Resident Memory (TRM) T-cells

```{r eval=FALSE}
umap = Seurat::FeaturePlot(Tcells_07, reduction='umap', features=c('rna_ZNF683'), label=TRUE, ncol=1) +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

vln = Seurat::VlnPlot(Tcells_07, c('rna_ZNF683'), ncol=1) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none")

umap | vln
```

```{r fig.width=10, fig.height=5, out.width='80%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.align='center'}
x = c()
y = c()
for(cluster in levels(tcells_07_meta$integrated_snn_res.0.1)){
  x = c(x, mean(tcells_07_meta$UMAP_1[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
  y = c(y, mean(tcells_07_meta$UMAP_2[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
}

umaps = list()
vlns = list()

for(gene in c('rna_ZNF683')){
  umaps[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=gene)) +
    ggplot2::geom_point(size=.6) + ggplot2::ggtitle(gsub('rna_', '', gene)) +
    ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2),
                                   limits=c(1e-5, NA)) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_07_meta$integrated_snn_res.0.1), color='black')
  
  vlns[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='integrated_snn_res.0.1', y=gene,
                                                                   fill='integrated_snn_res.0.1')) +
    ggplot2::geom_violin(trim=FALSE, scale='width') + ggplot2::theme_minimal() + ggplot2::theme(legend.position="none") +
    ggplot2::labs(x='', y='', title=gsub('rna_', '', gene)) + ggplot2::ylim(c(0,max(tcells_07_meta[, gene])))
}

final_plots = list()
for(gene in c('rna_ZNF683')){
  final_plots[[paste('a', gene, sep='_')]] = umaps[[gene]]
  final_plots[[paste('b', gene, sep='_')]] = vlns[[gene]]
}
patchwork::wrap_plots(final_plots, ncol=2)
```

#### Effector Memory (EM) T-cells

```{r eval=FALSE}
umap = Seurat::FeaturePlot(Tcells_07, reduction='umap', features=c('rna_KLRB1', 'rna_S100A4', 'rna_PTPRCAP', 'rna_PFN1', 'rna_GZMA'), label=TRUE, ncol=1) +
  ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2), limits=c(1e-5, NA)) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom")

vln = Seurat::VlnPlot(Tcells_07, c('rna_KLRB1', 'rna_S100A4', 'rna_PTPRCAP', 'rna_PFN1', 'rna_GZMA'), ncol=1) +
  ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none")

umap | vln
```

```{r fig.width=10, fig.height=10, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.align='center'}
x = c()
y = c()
for(cluster in levels(tcells_07_meta$integrated_snn_res.0.1)){
  x = c(x, mean(tcells_07_meta$UMAP_1[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
  y = c(y, mean(tcells_07_meta$UMAP_2[as.character(tcells_07_meta$integrated_snn_res.0.1)==cluster]))
}

umaps = list()
vlns = list()

for(gene in c('rna_KLRB1', 'rna_S100A4', 'rna_PTPRCAP', 'rna_PFN1', 'rna_GZMA')){
  umaps[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=gene)) +
    ggplot2::geom_point(size=.6) + ggplot2::ggtitle(gsub('rna_', '', gene)) +
    ggplot2::scale_color_gradientn(colours=c("navy", "yellow"), na.value=grDevices::rgb(.75, .75, .75, alpha=.2),
                                   limits=c(1e-5, NA)) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "bottom") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_07_meta$integrated_snn_res.0.1), color='black')
  
  vlns[[gene]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='integrated_snn_res.0.1', y=gene,
                                                                   fill='integrated_snn_res.0.1')) +
    ggplot2::geom_violin(trim=FALSE, scale='width') + ggplot2::theme_minimal() + ggplot2::theme(legend.position="none") +
    ggplot2::labs(x='', y='', title=gsub('rna_', '', gene)) + ggplot2::ylim(c(0,max(tcells_07_meta[, gene])))
}

final_plots = list()
for(gene in c('rna_KLRB1', 'rna_S100A4', 'rna_PTPRCAP', 'rna_PFN1', 'rna_GZMA')){
  final_plots[[paste('a', gene, sep='_')]] = umaps[[gene]]
  final_plots[[paste('b', gene, sep='_')]] = vlns[[gene]]
}
patchwork::wrap_plots(final_plots, ncol=4)
```


### Final Annotation

Looking into the expression of genes from the previous point, we knew that the cell-types of the clusters were the following:

| Clusters | Final_Annotation                   |
|:-------  |:---------------------------------- |
| 0        | Effector Memory (EM) CD4+ T-cells  |
| 1        | Naive CD4+ T-cells                 |
| 2        | Central Memory (CM) CD4+ T-cells   |

The annotations were stored in the metadata of the seurat object, with the variable names corresponding to the column in the previous table:

```{r eval=FALSE}
Tcells_07[['Final_Annotation']] = rep('', length(Tcells_07$integrated_snn_res.0.1))

EM_cells = rownames(Tcells_07@meta.data)[Tcells_07$integrated_snn_res.0.1=='0']
Tcells_07@meta.data[EM_cells, 'Final_Annotation'] = 'EM CD4'

naive_cells = rownames(Tcells_07@meta.data)[Tcells_07$integrated_snn_res.0.1=='1']
Tcells_07@meta.data[naive_cells, 'Final_Annotation'] = 'Naive CD4'

CM_cells = rownames(Tcells_07@meta.data)[Tcells_07$integrated_snn_res.0.1=='2']
Tcells_07@meta.data[CM_cells, 'Final_Annotation'] = 'CM CD4'
```

Then, we checked the previous SIGMA results and coloured clusters 0 and 7 by the new annotations:

```{r eval=FALSE}
cluster_0_cells = rownames(Tcells_07@meta.data)[Tcells_07$previous_0_7=='0']
cluster_7_cells = rownames(Tcells_07@meta.data)[Tcells_07$previous_0_7=='7']
subclusters_0 = Tcells_07$Final_Annotation[intersect(colnames(SIGMA_all_res02$input_parameters$expr), cluster_0_cells)]
subclusters_7 = Tcells_07$Final_Annotation[intersect(colnames(SIGMA_all_res02$input_parameters$expr), cluster_7_cells)]

SIGMA::plot_singular_vectors(SIGMA_all_res02, '0', colour=as.character(subclusters_0))
SIGMA::plot_singular_vectors(SIGMA_all_res02, '7', colour=as.character(subclusters_7))
```

```{r results='hide', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, include=FALSE}
cluster_0_cells = rownames(tcells_07_meta)[tcells_07_meta$previous_0_7=='0']
cluster_7_cells = rownames(tcells_07_meta)[tcells_07_meta$previous_0_7=='7']
subclusters_0 = tcells_07_meta[intersect(colnames(SIGMA_all_res02$input_parameters$expr), cluster_0_cells), 'Final_Annotation']
subclusters_7 = tcells_07_meta[intersect(colnames(SIGMA_all_res02$input_parameters$expr), cluster_7_cells), 'Final_Annotation']

sigma_C0 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '0', colour=as.character(subclusters_0))
sigma_C7 = SIGMA::plot_singular_vectors(SIGMA_all_res02, '7', colour=as.character(subclusters_7))

sigma_plots = patchwork::wrap_plots(list(sigma_C0, sigma_C7), ncol=2)
```

```{r fig.width=15, fig.height=5, out.width='100%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
sigma_plots
```

UMAP visualizations with cells coloured by the new annotations:

```{r eval=FALSE}
Seurat::DimPlot(Tcells_07, group.by='Final_Annotation', pt.size=0.5, label=T, label.size=3) +
  ggplot2::ggtitle('') + ggplot2::theme_minimal() + Seurat::NoLegend()
```

```{r fig.width=8, fig.height=5, out.width='80%', warning=FALSE, error=FALSE, message=FALSE, echo=FALSE, fig.align='center'}
tcells_anots = list()
for(level in c('Final_Annotation')){
  tcells_07_meta[,level] = factor(as.character(tcells_07_meta[,level]), levels=unique(tcells_07_meta[,level]))
  x = c()
  y = c()
  for(cluster in levels(tcells_07_meta[,level])){
    x = c(x, mean(tcells_07_meta$UMAP_1[as.character(tcells_07_meta[,level])==cluster]))
    y = c(y, mean(tcells_07_meta$UMAP_2[as.character(tcells_07_meta[,level])==cluster]))
  }
  tcells_anots[[level]] = ggplot2::ggplot(tcells_07_meta, ggplot2::aes_string(x='UMAP_1', y='UMAP_2', color=level)) +
    ggplot2::geom_point(size=.2) + ggplot2::ggtitle(level) +
    ggplot2::theme_minimal() + ggplot2::theme(legend.position = "none") +
    ggplot2::annotate(geom='text', x=x, y=y, label=levels(tcells_07_meta[,level]), color='black')
}
patchwork::wrap_plots(tcells_anots, ncol=1)
```


### Save Data

Finally, we saved the object. It was necessary for when we later on joined together the annotations from all clusters.

```{r eval=FALSE}
SeuratDisk::SaveH5Seurat(Tcells_07, paste(project_dir, '2_annotation/results_Tcells/datasets/Tcells_07.h5Seurat', sep='/'), overwrite=TRUE)
```


## Annotate Cluster 1

...



